Accessing the COVID-19 Death Rate in the US: While It Varies Across Different Income Levels, the Impact of Political Preference is Unignorable

### 2023-11-02 outputs/paper/paper.qmd
Update the Discussion
Second Draft

library(tidyverse)
library(here)
library(modelsummary)
library(gridExtra)
library(usmap)
library(knitr)
library(kableExtra)
library(purrr)

@@

All data was collected and cleaned using the statistical programming language `R` [@R]. The packages used are `tidyverse` [@tidyverse], `dplyr` [@dplyr], `stringr` [@stringr], and `janitor` [@janitor]. Besides, `here` [@here], `modelsummary` [@modelsummary], `girdExtra` [@gridExtra], `usmap` [@usmap], `knitr` [@knitr], `kableExtra` [@kableExtra] are used to analyze the data.

@@

```{r}
#| include: false
library(ggplot2)
library(maps)
states <- map_data("state")
counties <- map_data("county")
state_map <- data.frame(state.name = tolower(state.name), state.abb = state.abb)
counties$region <- state_map$state.abb[match(counties$region, state_map$state.name)]
counties$subregion <- tools::toTitleCase(counties$subregion)

mer <- merge(counties, data, by.x=c("subregion", "region"), by.y=c("county", "state"), all.x = TRUE)
mer$death_rate[is.na(mer$death_rate)] <- 0

ggplot(data=mer, mapping=aes(x=long, y=lat, group=group, fill=death_rate)) + 
  geom_polygon(color="black") +
  scale_fill_gradient(low="white", high="blue", name="Income") + 
  ggtitle('U.S. Map with Income by County') + 
  theme_minimal() + 
  theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank(),
        axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank()) +
  coord_fixed(1.3)

ggplot(mer, aes(x=long, y=lat, group=group)) + 
  geom_polygon(color="black", fill="gray") +
  coord_map() +
  theme_minimal()

duplicated_rows <- data[duplicated(data[c("county", "state")]),]
if (nrow(duplicated_rows) > 0) {
  print("There are duplicated rows!")
  print(duplicated_rows)
}
```

@@

## Politic preference to income levels: wealthier people are less likely to vote for the Republican, perhaps they are more educated.

@@

This is a surprising but also horrible observation as Donald Trump is especially in favor of poor people. Eight out of the top 10 counties with highest poverty rate voted for Donald Trump. Referring to @fig-map, we can clearly see that the above 80% states in the poorer states, middle US, had voted for Trump. Conversely, above 80% wealthier coastal states, such as California and New York, had voted for The Democratic party. These both verify our observations from the model. What Trump's votes does not know is that their support may bring more deaths to their counties.

But why, why the counties with higher income levels are less likely to vote for the Republican party and hence results in a lower death rate? One thing that can explain this pattern from our model is the proportion of people that have higher education level. According to @tbl-second_model, we can find that there is an inverse relationship between the death rate and the number of residents with high education level. Its value of coefficient is much higher than the previous two, indicating that education level is an essential factor in predicting the death rate.

@@

Therefore, the death rate increases as people voting for the Republican increases. However, people living at poorer counties are more likely to vote for Trump and one potential reason behind this is that the the proportion of residents with at least bachelor degree is relatively low in those counties. High education level people are more willing to live at the wealthier counties, and they are less likely to vote for Trump. Perhaps more educated people do not satisfy the actions taken by the government.

## Income levels to Health insurance: more people can afford the private insurance in wealthier counties and reduce the deaths rate.
In addition to the political preferences, the healthcare insurance is also significant in predicting the death rate for each county. Based on @tbl-second_model, it seems that both `no_insurance` and `private_insurance` are also significant in predicting the death rate for each county but they shows an inverse relationship on the death rate. In addition, the values of these two coefficients are almost same, meaning that the effects on death rate between them can be offset.

@@

@@

# References

@@

>>>>

library(tidyverse)
library(here)
library(modelsummary)
library(gridExtra)
library(usmap)
library(knitr)
library(kableExtra)
library(purrr)
library(ggplot2)

@@

All data was collected and cleaned using the statistical programming language `R` [@R]. The packages used are `tidyverse` [@tidyverse], `dplyr` [@dplyr], `stringr` [@stringr], and `janitor` [@janitor]. Besides, `here` [@here], `modelsummary` [@modelsummary], `girdExtra` [@gridExtra], `usmap` [@usmap], `ggpot2` [@ggplot2], `knitr` [@knitr], `kableExtra` [@kableExtra] are used to analyze the data.

@@

@@

## Although income levels are known to affect the COVID mortality, the impact of political preferences is more substantial than anticipated, potentially overcshadowing the effects of income.

@@

```{r, fig.width=10, fig.height=4}
#| echo: false
#| message: false
#| label: fig-income_state
#| fig-cap: Variation in COVID-19 Mortality Rates Across States, Categorized by Income Levels and Dominant Political Preferences

library(dplyr)
library(ggplot2)
# Find the wining data for each state
data_win_party <- data %>%
  group_by(state) %>%
  summarise(dem_votes = sum(votes_dem),
            rep_votes = sum(votes_rep), 
            .groups = 'drop') %>% 
  mutate(winning_party = ifelse(dem_votes > rep_votes, 'Democratic', 'Republican')) %>% 
  select(state, winning_party)
# Calculate the average income
average_income_by_state <- data %>%
  group_by(state) %>%
  summarise(average_income = mean(mean_household_income), .groups = 'drop')
# Join the winning party and average income back to the original data
data_win_party_income <- data %>%
  left_join(data_win_party, by = "state") %>%
  left_join(average_income_by_state, by = "state")
# Arrange states by average income
ordered_states <- data_win_party_income %>%
  arrange(average_income) %>%
  .$state %>%
  unique()
# Create a factor with levels based on income-ordered states for plotting
data_win_party_income$state <- factor(data_win_party_income$state, levels = ordered_states)

average_death_rate <- mean(data_win_party_income$death_rate)

ggplot(data_win_party_income, aes(x = state, y = death_rate, fill = winning_party)) +
  geom_boxplot() +
  geom_hline(yintercept = average_death_rate, linetype = "dashed", color = "black") +
  annotate("text", x = Inf, y = average_death_rate, label = " National Average", hjust = 1, vjust = -0.5, color = "black", size = 3) +

  # Adding the dashed line
  scale_fill_manual(values = c('Democratic' = 'blue', 'Republican' = 'red')) +
  labs(y = 'Death Rate', x = 'State (Ascending Income)', fill = "Winning Party") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

@fig-income_state can prove our point. The @fig-income_state shows the spread of COVID death rate for each state in the order of ascend income levels. The color represent the party each state supported during 2020 US Federal Election. We can observe that twelve out of the top 15 poorest counties voted for Donald Trump. In contrast, only one of the top 15 richest states voted for the Republican Party. In terms of the death rate, the @fig-income_state also shows that the death rate decreases as income level increases and majority of counties voting for Democratic party has a death rate lower than the national average level.

```{r}
#| echo: false
#| message: false
library(dplyr)
library(ggplot2)

# Assuming 'data' is your DataFrame with columns: pctile, rep_rate, dem_rate

# Step 1: Calculate the average support rate for each party at each income percentile
avg_support <- data %>%
  group_by(pctile) %>%
  summarise(
    avg_rep_rate = mean(rep_rate),
    avg_dem_rate = mean(dem_rate)
  ) %>%
  ungroup()

# Step 2: Divide income percentile into four groups
avg_support <- avg_support %>%
  mutate(
    pctile_group = cut(pctile, breaks = quantile(pctile, probs = 0:4/4), include.lowest = TRUE)
  )

# Step 3: Fit regression models and predict values for each piece (this is a placeholder)
# You would fit your piece-wise regression models here for each party and each group

# Plotting with placeholder regression lines (You'll need the actual fitted model predictions)
ggplot(avg_support, aes(x = pctile)) +
  geom_point(aes(y = avg_rep_rate), color = 'red') +
  geom_point(aes(y = avg_dem_rate), color = 'blue') +
  # Placeholder for regression lines
  geom_smooth(data = subset(avg_support, pctile_group == 1), aes(y = avg_rep_rate, color = "Republican"), method = "lm", se = FALSE) +
  geom_smooth(data = subset(avg_support, pctile_group == 1), aes(y = avg_dem_rate, color = "Democrat"), method = "lm", se = FALSE) +
  # ... repeat geom_smooth for other groups ...
  scale_color_manual(values = c("Republican" = "red", "Democrat" = "blue")) +
  labs(title = 'Average Support Rate by Income Percentile',
       x = 'Income Percentile', y = 'Average Support Rate') +
  theme_minimal() +
  theme(legend.position = "bottom")

# Note: In a real scenario, you would repeat the geom_smooth for the other pctile_groups (2, 3, 4)
# and would use the actual fitted model predictions.

```

This is a surprising but also horrible observation as Donald Trump is especially in favor of poor people. This means the low-income voters voting for Trump increase their chances of death in disguised form. Based on the fact that the death of COVID of low-income people is almost two times higher than the high-income [citation], it seems that it is not unfair to attribute the high death rate of US to Republican Party (Donald Trump).

## The difference in high-education atteinment and insurance structure may be potential factors causing the above phenomenon.

The model [@tbl-second_model] incorporates the proportion of people that have at least bachelor degree, and the proportion of people with private insurance or without insurance. In terms of education level, there is an inverse relationship between the death rate and the number of residents with high education level. Its coefficient value is relatively high, the death number can be decreased by 0.24 if there are ten more percent people having higher degree.

@@

Therefore, one reason can explain the inequality of COVID mortality is that there are less people with higher degree lives in the low-income county. High educational level people are more willing to live at the wealthier counties. They are more educated and perhaps they know what Trump did was wrong and hence they did not vote for Trump. This can also explain why there are less votes for Trump in wealthier counties/states.

In addition to education level, the healthcare insurance is also significant in predicting the death rate for each county. Based on @tbl-second_model, it seems that both `no_insurance` and `private_insurance` are also significant in predicting the death rate for each county but they shows an inverse relationship on the death rate. In addition, the values of these two coefficients are almost same, meaning that the effects on death rate between them can be offset.

@@

Comparing education with insurance, one common thing is that the we can see diverging trends in poorer and wealthier counties. People living in wealthier counties are more educated and concerns more about the public issue. Meanwhile, their awareness of protecting themselves from COVID is stronger than the poorer people. Due to their wealth level, they have the money to purchase private insurance and are more accessible to the medicines or vaccines. This explains why the wealthier counties had a relatively lower COVID mortality rate than poor ones.

@@

# References {.unnumbered}

@@

https://www.reuters.com/world/us/us-poor-died-much-higher-rate-covid-than-rich-report-2022-04-04/
